### 2012-10-17 rnaseq_paper.tex

@@

>>>>

We next used Tophat to align mRNAseq reads to the genome and partition
reads by chromosome; we then assembled the partitioned reads with
Velvet and Oases using the same range of parameters as the global
assembly, above.  While the local assemblies were considerably more
computationally efficient, they lacked several thousand unique regions
that were present in the global assembly
(Table~\ref{unique_sequences_matched_mouse}); this is probably due to
the incomplete nature of the current chicken genome (cite).
Interestingly, over a hundred regions were present {\em only} in the
local assemblies, suggesting that the local assemblies might be
recovering additional exons.  Significant numbers of unique regions
from both global and local assemblies showed homology to the mouse
genome, indicating that at least some of these unique sequences
represented real sequence content.  Figure~\ref{global_vs_local} shows
an example of different isoforms detected by the two assembly methods.

We developed an exon graph approach to construct gene models from
alignments of transcripts against the genome.  Our approach, called
Gimme, uses exon graphs to merge transcripts and gene models based on
overlapping exons (see Materials and Methods).  We used Gimme to
obtain 20,439 gene models containing 25,859 isoforms from our global
assembly; 21,963 gene models containing 25,859 XX from our local
assembly; and 20,922 gene models containing 27,732 isoforms from the
merged global and local assemblies (Table~\ref{genes_transcripts}).

(more here - what? @CTB)

\subsubsection{Most splice junctions are independently supported}

As shown above, Cufflinks and the assembly method detect a number of
distinct but equally valid splice junctions, which suggests that we
could obtain greater sensitivity by merging both sets of predictions.
We therefore used Gimme to merge the Cufflinks and assembly gene
models Table.~\ref{genes_transcripts}.  This resulted in a decreased
number of total genes, suggesting that some fragmented genes were
merged together to form more complete gene structures (e.g. see
Fig.~\ref{cufflinks_asm}).
% @CTB what about splice junctions etc??

\subsection{Leftover text}

The program can merge all transcripts from multiple-kmers and predict a structure of full-length isoforms and genes.
As shown in figure~\ref{model_comparisons}, our gene models are comparable to other existing gene models.
Total of 14,703 genes were obtained from global assembly and 14,931 genes from local assembly.
However, the number of genes obtained from global + local assembly is only 15,411.
The number of genes and isoforms slightly increased after combining transcripts from global and local
assembly together (Table~\ref{genes_transcripts}). 
This indicates that some transcripts from both methods were merged together.

Most transcripts include a large part of untranslated regions (UTRs), especially $3'$ UTR in our datasets.
These regions are challenging to predict correctly solely from computational methods due to low degree of conservation [?].
RNA-Seq-derived gene models are useful for studying variations within UTR regions, which may be involved in regulation of isoform expression\cite{} (Fig.~\ref{long_utr}).

The number of genes from our pipeline might be overestimated due to
fragmented transcripts.  In our datasets, read coverage is lower at
the $5'$ end due to the method used to convert mRNA to cDNA in library
preparation.  A bias of read coverage toward $3'$ end and low
expression level result in fragmented transcripts as shown in
Fig.~\ref{fragmented_transcripts}.  Moreover, spurious splice
junctions due to duplication in genome sequence can lead to a large
number of dubious isoforms.  This seems to occur in a few genome
sequences such as chromosome E64\_random.
Figure~\ref{isoforms_boxplot} shows that a majority of genes from our
gene models have only a few isoforms.

In addition, parameters used in building gene models such as minimum
exon/intron size determine how transcripts are merged together, which
can increases or decreases the number of isoforms.  In this study, we
adjust the parameters to enhance sensitivity of splice junctions
detection that may result in overestimation of the number of isoforms.
=======

@@

\subsubsection{Cufflinks}

To evaluate an efficiency of our pipeline in detecting splice junctions,
we compared the number of splice junctions detected by our pipeline to that from , ESTs and
Ensembl gene models version 64.
Cufflinks detected 89,289 (86.49\%) splice junctions that are also detected by our pipeline.
However, 37.61\% (5,565) of junctions not detected by Cufflinks are supported by ESTs.
This suggests that the assembly detects some splice junctions that are not detected by Cufflinks and they are
genuine splice junctions.

We also aligned transcripts from Cufflinks and the pipeline to Ensembl transcripts to compare genes and transcripts detected by
both methods.
We selected only transcripts that match with more than 90\% identity and cover more than 80\% of an Ensembl transcript.
Our pipeline detects 7,046 genes and Cufflinks detects 8,341 genes of total 17,934 Ensembl genes.
6,572 of those genes are both detectd by our pipeline and Cufflinks.
In addition, we wanted to compare the number of isoforms detected by our pipeline and Cufflinks within the same gene; however,
a transcript can match to multiple isoforms in the same gene, especially when it is not complete.
Therefore, we compare the number of splice junctions detected by both methods within the same gene instead.
From 6,572 genes, our pipeline detects 57,437 splice junctions, whereas Cufflinks detects 62,101 junctions.

However, because Ensembl gene models do not include all isoforms or splice junctions from ESTs, we investigated splice junctions
that are not in Ensembl but are supported by ESTs in those 6,572 genes and found that our pipeline detects 2,260 splice junctions
and Cufflinks only detects 1,588 splice junctions. Figure ~\ref{alt_splice_site} shows an example of splice junctions not included in Ensembl gene models, which are detected by our pipeline.

To conclude, Cufflinks performs better than our pipeline for detecting both genes and splice junctions in Ensembl gene models.
However, we found that our pipeline detects more splice junctions that are not in Ensembl but are supported by ESTs.
=======
The results show that our pipeline predicts many gene models and
isoforms after assembly.  We validated these gene models in several
different ways.
